---
title: "Current and future range maps of mammals made by biomodelos"
description: |
  Quick review
author:
  - name: Lei Song
    url: https://github.com/LLeiSong
    affiliation: University of California, Santa Barbara
output: 
  distill::distill_article:
    toc: true
    toc_depth: '2'
header-includes: 
  \usepackage{caption}
  \renewcommand{\figurename}{Fig.} 
  \renewcommand{\thefigure}{S\arabic{figure}}
  \renewcommand{\thetable}{S\arabic{table}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(ggplot2)
library(terra)
library(sf)
library(dplyr)
library(ggpubr)
library(tidyterra)
terraOptions(progress=0)

# Set directories and paths
data_dir <- here("data/mamiferos")
species_list <- list.files(data_dir)
cur_path <- "ensembles/current/MAXENT/%s_10_MAXENT.tif"
fut_path <- "ensembles/future/MAXENT/future_ensemble/2021-2040_ssp126_10.tif"
aoi_path <- "interest_areas/shape_M.shp"

col <- st_read(here("data/colombia.geojson"))
fname <- here("data/sp_numbers.csv")
sp_numbers <- read.csv(fname)
```

## Some example species

To keep the explanation straightforward, we only included the comparison between the current range (_species_10_MAXENT.tif_) and the future range under SSP126 (2011–2040) (_2021-2040_ssp126_10.tif_), using a 10% threshold. 

Here, we compared four representative species: Lucifer titi monkey (_Cheracebus lucifer_), Nancy Ma's night monkey (_Aotus nancymaae_), Varied white-fronted capuchin (_Cebus versicolor_), and Ornate titi monkey (_Plecturocebus ornatus_).

The following two species has dramatic **range shift**.

### Lucifer titi monkey (_Cheracebus lucifer_)

```{r, echo = FALSE, include = TRUE, warning=FALSE, message=FALSE}
sp <- "Cheracebus.lucifer"

cur_fname <- file.path(data_dir, sp, sprintf(cur_path, gsub("\\.", "_", sp)))
fut_fname <- file.path(data_dir, sp, fut_path)
aoi_fname <- file.path(data_dir, sp, aoi_path)

cur_range <- rast(cur_fname) %>% extend(col)
aoi <- st_read(aoi_fname, quiet = TRUE) %>% st_union()
fut_range <- rast(fut_fname) %>% extend(col)

g1 <- ggplot() +
  geom_spatraster(data = cur_range, na.rm = TRUE) +
  geom_sf(data = aoi, fill = "transparent", color = "blue") +
  geom_sf(data = col, fill = "transparent", color = "black") +
  scale_fill_whitebox_c(
    name = "Range",
    palette = "muted",
    n.breaks = 2,
    guide = guide_legend(reverse = TRUE)) +
  theme_void()

g2 <- ggplot() +
  geom_spatraster(data = fut_range, na.rm = TRUE) +
  geom_sf(data = aoi, fill = "transparent", color = "blue") +
  geom_sf(data = col, fill = "transparent", color = "black") +
  scale_fill_whitebox_c(
    name = "Range",
    palette = "muted",
    n.breaks = 2,
    guide = guide_legend(reverse = TRUE)) +
  theme_void()

ggarrange(g1, g2, labels = c("Current", "SSP126 2011-2040"))
```

### Nancy Ma's night monkey (_Aotus nancymaae_)

```{r, echo = FALSE, include = TRUE, warning=FALSE, message=FALSE}
sp <- "Aotus.nancymaae"

cur_fname <- file.path(data_dir, sp, sprintf(cur_path, gsub("\\.", "_", sp)))
fut_fname <- file.path(data_dir, sp, fut_path)
aoi_fname <- file.path(data_dir, sp, aoi_path)

cur_range <- rast(cur_fname) %>% extend(col)
aoi <- st_read(aoi_fname, quiet = TRUE) %>% st_union()
fut_range <- rast(fut_fname) %>% extend(col)

g1 <- ggplot() +
  geom_spatraster(data = cur_range, na.rm = TRUE) +
  geom_sf(data = aoi, fill = "transparent", color = "blue") +
  geom_sf(data = col, fill = "transparent", color = "black") +
  scale_fill_whitebox_c(
    name = "Range",
    palette = "muted",
    n.breaks = 2,
    guide = guide_legend(reverse = TRUE)) +
  theme_void()

g2 <- ggplot() +
  geom_spatraster(data = fut_range, na.rm = TRUE) +
  geom_sf(data = aoi, fill = "transparent", color = "blue") +
  geom_sf(data = col, fill = "transparent", color = "black") +
  scale_fill_whitebox_c(
    name = "Range",
    palette = "muted",
    n.breaks = 2,
    guide = guide_legend(reverse = TRUE)) +
  theme_void()

ggarrange(g1, g2, labels = c("Current", "SSP126 2011-2040"))
```

The following two species have dramatic **range shrink**.

### Varied white-fronted capuchin (_Cebus versicolor_)

```{r, echo = FALSE, include = TRUE, warning=FALSE, message=FALSE}
sp <- "Cebus.versicolor"

cur_fname <- file.path(data_dir, sp, sprintf(cur_path, gsub("\\.", "_", sp)))
fut_fname <- file.path(data_dir, sp, fut_path)
aoi_fname <- file.path(data_dir, sp, aoi_path)

cur_range <- rast(cur_fname) %>% extend(col)
aoi <- st_read(aoi_fname, quiet = TRUE) %>% st_union()
fut_range <- rast(fut_fname) %>% extend(col)

g1 <- ggplot() +
  geom_spatraster(data = cur_range, na.rm = TRUE) +
  geom_sf(data = aoi, fill = "transparent", color = "blue") +
  geom_sf(data = col, fill = "transparent", color = "black") +
  scale_fill_whitebox_c(
    name = "Range",
    palette = "muted",
    n.breaks = 2,
    guide = guide_legend(reverse = TRUE)) +
  theme_void()

g2 <- ggplot() +
  geom_spatraster(data = fut_range, na.rm = TRUE) +
  geom_sf(data = aoi, fill = "transparent", color = "blue") +
  geom_sf(data = col, fill = "transparent", color = "black") +
  scale_fill_whitebox_c(
    name = "Range",
    palette = "muted",
    n.breaks = 2,
    guide = guide_legend(reverse = TRUE)) +
  theme_void()

ggarrange(g1, g2, labels = c("Current", "SSP126 2011-2040"))
```

### Ornate titi monkey (_Plecturocebus ornatus_)

```{r, echo = FALSE, include = TRUE, warning=FALSE, message=FALSE}
sp <- "Plecturocebus.ornatus"

cur_fname <- file.path(data_dir, sp, sprintf(cur_path, gsub("\\.", "_", sp)))
fut_fname <- file.path(data_dir, sp, fut_path)
aoi_fname <- file.path(data_dir, sp, aoi_path)

cur_range <- rast(cur_fname) %>% extend(col)
aoi <- st_read(aoi_fname, quiet = TRUE) %>% st_union()
fut_range <- rast(fut_fname) %>% extend(col)

g1 <- ggplot() +
  geom_spatraster(data = cur_range, na.rm = TRUE) +
  geom_sf(data = aoi, fill = "transparent", color = "blue") +
  geom_sf(data = col, fill = "transparent", color = "black") +
  scale_fill_whitebox_c(
    name = "Range",
    palette = "muted",
    n.breaks = 2,
    guide = guide_legend(reverse = TRUE)) +
  theme_void()

g2 <- ggplot() +
  geom_spatraster(data = fut_range, na.rm = TRUE) +
  geom_sf(data = aoi, fill = "transparent", color = "blue") +
  geom_sf(data = col, fill = "transparent", color = "black") +
  scale_fill_whitebox_c(
    name = "Range",
    palette = "muted",
    n.breaks = 2,
    guide = guide_legend(reverse = TRUE)) +
  theme_void()

ggarrange(g1, g2, labels = c("Current", "SSP126 2011-2040"))
```

## Spatially constraint for future ranges

After reviewing these layers, we observed that future ranges often extend beyond current ranges very much, which may not be realistic. To address this, we proposed limiting future ranges to a meaningful buffer around current ranges. Ideally, this buffer should be set individually for each species based on their dispersal ability. However, in the absence of such information, we applied a 100 km buffer to the **interested area**.

With this constraint in place, we calculated the Intersection over Union (IoU), which is determined by dividing the area of intersection by the area of union.

Below is the IoU distribution, which appears reasonable.

```{r, echo = FALSE, include = TRUE, warning=FALSE, message=FALSE}
ggplot(sp_numbers, aes(x = IoU)) +
  geom_histogram(aes(y = ..density..), binwidth = 5, 
                 colour = "black", fill = "white") +
  geom_density(alpha = .2, fill="#FF6666") +
  ylab("Density") +
  theme_pubclean(base_size = 12) +
  theme(axis.text = element_text(size = 12, color = "black"))
```

Additionally, the mammal richness calculated with this constraint are as follows:

```{r, echo = FALSE, include = TRUE, warning=FALSE, message=FALSE}
# Examine the overall current and future distribution
sp_selected <- sp_numbers %>% filter(IoU <= 50 & aoi_area >= 1e6)
col_ext <- vect(col) %>% ext()

cur_diversity <- lapply(sp_selected$species, function(sp){
  cur_fname <- file.path(data_dir, sp, sprintf(cur_path, gsub("\\.", "_", sp)))
  rast(cur_fname) %>% extend(col_ext)
})
cur_diversity <- do.call(c, cur_diversity)
cur_diversity <- sum(cur_diversity, na.rm = TRUE)

fut_diversity <- lapply(sp_selected$species, function(sp){
  fut_fname <- file.path(data_dir, sp, fut_path)
  aoi_fname <- file.path(data_dir, sp, aoi_path)
  
  aoi <- st_read(aoi_fname, quiet = TRUE)
  aoi_spread <- st_buffer(aoi, 0.008333333 * 100)
  
  mask(rast(fut_fname), aoi_spread) %>% extend(col_ext)
})
fut_diversity <- do.call(c, fut_diversity)
fut_diversity <- sum(fut_diversity, na.rm = TRUE)

diversity <- c(cur_diversity, fut_diversity)
diversity <- mask(diversity, col) %>% trim()
names(diversity) <- c("Current", "2021-2040_ssp126")

ggplot() +
  geom_spatraster(data = diversity) +
  scale_fill_whitebox_c(
    name = "Diversity",
    palette = "muted",
    n.breaks = 8,
    guide = guide_legend(reverse = TRUE)) + 
  facet_wrap(~lyr) + 
  theme_void() +
  theme(title = element_text(size = 14),
        legend.text = element_text(size = 14),
        strip.text = element_text(size = 14))
```

## Overall distribution of range areas

Finally, here is the overall review of the species ranges (I haven't removed flying mammals yet, but I may do so):

```{r, echo = FALSE, include = TRUE, warning=FALSE, message=FALSE, fig.height=7, fig.width=6}
# Summarize the species
nums <- c(
  (sp_numbers %>% filter(aoi_area < 1e6) %>% nrow()) / nrow(sp_numbers) * 100,
  (sp_numbers %>% filter(aoi_area >= 1e6 & IoU > 50) %>% nrow()) / 
    nrow(sp_numbers) * 100,
  (sp_numbers %>% filter(aoi_area >= 1e6 & IoU <= 50) %>% nrow()) / 
    nrow(sp_numbers) * 100)
names(nums) <- c("small", "large_health", "large_fragment")

# Make a figure
ggplot(data = sp_numbers) +
  geom_vline(xintercept = 1, color = "purple", linetype = "dashed") + 
  geom_segment(aes(x = 1, xend = 10, y = 50, yend = 50), 
               color = "purple", linetype = "dashed") +
  geom_point(aes(x = aoi_area / 1e6, y = IoU, 
                 color = cur_area / 1e6, size = fut_area / 1e6)) +
  geom_smooth(aes(x = aoi_area / 1e6, y = IoU)) + xlim(0, 10) + 
  scale_color_viridis_c(expression(paste("Current area(1e"^6, ")")),
                        guide = guide_colorbar(title.vjust = 1)) +
  scale_size_area(expression(paste("Future area(1e"^6, ")"))) +
  xlab(expression(paste("Area of the interested area (1e"^6, ")"))) + ylab("IoU") +
  annotate("text", label = paste0(round(nums, 1), "%"), 
           x = c(0, 9, 9), y = c(90, 55, 12), size = 5, fontface = "bold") +
  theme_pubclean() +
  theme(legend.box = "vertical",
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12))
```
