---
title: "Climate adaptive habitat connectivity"
description: |
  Using biomodelos as the base layers
author:
  - name: Lei Song
    url: https://github.com/LLeiSong
    affiliation: University of California, Santa Barbara
output: 
  distill::distill_article:
    toc: true
    toc_depth: '2'
header-includes: 
  \usepackage{caption}
  \renewcommand{\figurename}{Fig.} 
  \renewcommand{\thefigure}{S\arabic{figure}}
  \renewcommand{\thetable}{S\arabic{table}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(ggplot2)
library(ggpubr)

root_dir <- here::here()
fname <- file.path(root_dir, "data/sp_numbers.csv")
dispersal_path <- file.path(
  root_dir, "data/dispersal", "species_dispersal_rate.csv")
data_dir <- file.path(root_dir, "data/mamiferos")
cur_path <- "ensembles/current/MAXENT/%s_10_MAXENT.tif"
fut_paths <- sprintf(
  "%s/2021-2040_ssp%s_10.tif", 
  "ensembles/future/MAXENT/future_ensemble", c("126", "585"))
aoi_path <- "interest_areas/shape_M.shp"
crs_analysis <- "ESRI:54030"
```

## Check the current and future ranges

```{r}
## Later reload the table from file
sp_numbers_all <- read.csv(fname)

# Load boundary
col_real <- st_read(file.path(root_dir, "data/colombia.geojson")) %>% 
  st_transform(crs_analysis)
col <- vect(col_real) %>% ext()

# Summarize the species
for (ssp in c("ssp126", "ssp585")){
  sp_numbers <- sp_numbers_all[sp_numbers_all$ssp == ssp, ]
  nums <- c(
    (sp_numbers %>% filter(aoi_area < 1e6) %>% nrow()) / nrow(sp_numbers) * 100,
    (sp_numbers %>% filter(aoi_area >= 1e6 & IoU > 50) %>% nrow()) / 
      nrow(sp_numbers) * 100,
    (sp_numbers %>% filter(aoi_area >= 1e6 & IoU <= 50) %>% nrow()) / 
      nrow(sp_numbers) * 100)
  names(nums) <- c("small", "large_health", "large_fragment")
  
  # Make a figure
  ggplot(data = sp_numbers) +
    geom_vline(xintercept = 1, color = "purple", linetype = "dashed") + 
    geom_segment(aes(x = 1, xend = 10, y = 50, yend = 50), 
                 color = "purple", linetype = "dashed") +
    geom_point(aes(x = aoi_area / 1e6, y = IoU, 
                   color = cur_area / 1e6, size = fut_area / 1e6)) +
    geom_smooth(aes(x = aoi_area / 1e6, y = IoU)) + xlim(0, 10) + 
    scale_color_viridis_c(expression(paste("Current area(1e"^6, ")")),
                          guide = guide_colorbar(title.vjust = 1)) +
    scale_size_area(expression(paste("Future area(1e"^6, ")"))) +
    xlab(expression(paste("AOI area (1e"^6, ")"))) + ylab("IoU") +
    annotate("text", label = paste0(round(nums, 0), "%"), 
             x = c(0, 9, 9), y = c(90, 55, 12), size = 5, fontface = "bold") +
    theme_pubclean() +
    theme(legend.box = "vertical",
          axis.text = element_text(size = 11),
          axis.title = element_text(size = 12))
  
  ggsave(sprintf("results/area_distribution_%s.png", ssp), 
         width = 6, height = 7, bg = "white")
  
  # Examine the overall current and future distribution
  sp_selected <- sp_numbers %>% filter(IoU <= 50 & aoi_area >= 1e6)
  
  cur_diversity <- lapply(sp_selected$species, function(sp){
    cur_fname <- file.path(data_dir, sp, sprintf(cur_path, gsub("\\.", "_", sp)))
    rast(cur_fname) %>% project(crs_analysis) %>% extend(col)
  })
  cur_diversity <- do.call(c, cur_diversity)
  cur_diversity <- sum(cur_diversity, na.rm = TRUE)
  
  fut_diversity <- lapply(sp_selected$species, function(sp){
    fut_path <- fut_paths[grep(ssp, fut_paths)]
    fut_fname <- file.path(data_dir, sp, fut_path)
    aoi_fname <- file.path(data_dir, sp, aoi_path)
    
    # Get potential dispersal distance
    dispersal_rate <- read.csv(dispersal_path) %>% 
      filter(species == gsub("\\.", " ", sp)) %>% pull(dispersal_rate)
    if (length(dispersal_rate) == 0) {
      dispersal_rate <- read.csv(dispersal_path) %>% 
        pull(dispersal_rate) %>% median()
    }
    dis_dist <- dispersal_rate * 40 * 1000 # in meter
    
    aoi <- st_read(aoi_fname) %>% st_transform(crs_analysis)
    aoi_spread <- st_buffer(aoi, dis_dist)
    
    mask(rast(fut_fname) %>% project(crs_analysis), aoi_spread) %>% extend(col)
  })
  fut_diversity <- do.call(c, fut_diversity)
  fut_diversity <- sum(fut_diversity, na.rm = TRUE)
  
  diversity <- c(cur_diversity, fut_diversity)
  diversity <- mask(diversity, col_real) %>% trim()
  names(diversity) <- c("Current", "2021-2040")
  
  ggplot() +
    geom_spatraster(data = diversity) +
    scale_fill_whitebox_c(
      name = "Diversity",
      palette = "muted",
      n.breaks = 8,
      guide = guide_legend(reverse = TRUE)) + 
    facet_wrap(~lyr) + 
    theme_void() +
    theme(title = element_text(size = 14),
          legend.text = element_text(size = 14),
          strip.text = element_text(size = 14))
  
  ggsave(sprintf("results/diversity_%s.png", ssp), 
         width = 8, height = 5, bg = "white")
}
```


## Set up the HPC environment

**Step1**. Clone the previous conda environment sdm, which is ready to use R for spatial analysis.

```
conda create --name conn --clone sdm
conda activate conn
```

**Step2**. Install Julia and Omniscape package

```
# Install Juliaup manager and julia
conda install -c conda-forge juliaup
juliaup add <version> # or juliaup add release
juliaup config channelsymlinks true

# Get the Julia path for R call later
# ~/anaconda3/envs/conn/bin/julia-1.11.3
which julia-1.11.3 # or which julia-release

# Install customized omniscape
using Pkg
]

# Extra steps because of the issue in ArchGDAL
add PROJ_jll@901.300 GDAL_jll@301.900
add ArchGDAL

# Link and compile Omniscape
dev /home/lsong/Omniscape.jl
using Omniscape

# Add other packages for R
add Suppressor
add RCall

# Log out of conda if necessary
conda deactivate
```

## Omniscape

A few terms:

**Source strength**: Similar to conductance but represents the number of individuals originating from a given pixel.

**Resistance layer**: The inverse of conductance, representing the difficulty of movement across the landscape.

**Target pixel**: The central pixel within the circular window.

**Source pixels**: All pixels within the circular window that exceed the defined source strength threshold. They are identified similarly to the target pixel.

Omniscape essentially operates Circuitscape in advanced mode, treating the central target pixel as the ground and the source pixels as current sources to model connectivity dynamics.
