---
title: "Connectivity in land use planning"
author: "Lei Song"
date: "2024-05-23"
output: html_document
---

<style type="text/css">
  body{
  font-size: 16pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(terra)
library(dplyr)
library(here)
library(scrubr)
library(ggplot2)
library(tidyterra)
library(ggpubr)

sdm_dir <- file.path(here("data/suitability"), "sdm")
species <- "Saguinus oedipus"
```

## Background

Took approaches conducted in paper [Hanson et. al, 2022](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2664.14251).

- **Total area**: Connectivity increase with increasing total habitat. This is an PA-level only approach. (We have already done better by using species-level. Also mathematically, it may against species targets and 30% target.)
- Species representation: This is a species-level "total area" approach. (May be duplicated with habitat targets.)
- Boundary length: To minimize the total boundary (perimeter) length of reserves. (Good luck if you insist on this.)
- Human pressure: To aggressively select adjacent planning units with low human pressure.
- Naturalness-based landscape resistance: To aggressively select adjacent planning units with natural conditions. (This seems like the opposite of "Human pressure".)
- Focal species landscape resistance: To select adjacent planning units that can strongly facilitate movements of a single (or a few) focal species.
- Habitat heterogeneity: To select adjacent planning units with **low** habitat heterogeneity. *This assumes that places with greater habitat heterogeneity have reduced species' movements* (I am not a big fan of this theory.).
- Environmental similarity: To select adjacent planning units with similar environmental conditions. I think Environmental similarity is another word for species suitability. *This can potentially promote the consideration of climate change.*

So, we can try "naturalness-based landscape resistance", "focal species landscape resistance", or "Environmental similarity" (species suitability).

## Experiments

```{r, eval=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Range map
range_map <- rast(
  file.path("data/IUCN_AOH_100m/mammals",
            sprintf("%s.tif", gsub(" ", "_", species))))
range_map[range_map == 0] <- NA

# Calculate the distance
dist_to_aoh <- distance(range_map, unit = "km")
dist_to_aoh <- mask(dist_to_aoh, suit)
dist_to_aoh <- (dist_to_aoh - global(dist_to_aoh, "min", na.rm = TRUE)[[1]]) / 
  (global(dist_to_aoh, "max", na.rm = TRUE)[[1]] - 
     global(dist_to_aoh, "min", na.rm = TRUE)[[1]])
writeRaster(dist_to_aoh, "data/suitability/dist_to_aoh.tif")
```

## Use zonation5

We can use zonation5 to pre-process species layers to consider both climate change and connectivity before prioritizer. 

What described in zonation manual makes perfect sense:

1. Are the different layers (consecutive) time-steps or alternative realizations? I believe they mean alternative realizations to us.

2. Are the different time-steps/realizations equally likely? Definitely not. The future is inherently more uncertain than the present.

3. Is there a need to account for connectivity? We want! But we have to deal time-steps as time-steps.

*4. Priority areas produced in such analyses cannot necessarily be interpreted as ecological refugia.* This is completely right. Therefore, we will use the rankmaps as the inputs for prioritizer. 

```{r, eval=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Prepare layers, current and future (alternative realizations)
# Current layer
fn <- file.path(sdm_dir, sprintf("suit_%s.tif", gsub(" ", "_", species)))
suit <- rast(fn) %>% mean(na.rm = TRUE)
suit[suit < 0] <- 0

fn <- file.path(sdm_dir, sprintf("future_suit_%s.tif", gsub(" ", "_", species)))
future_suit <- rast(fn) %>% mean(na.rm = TRUE)
future_suit[future_suit < 0] <- 0

pas <- rast("mod_input_lyr/nps.tif")

suit <- project(suit, pas, method = "bilinear")
future_suit <- project(future_suit, pas, method = "bilinear")

writeRaster(suit, "data/zonation/data/suit.tif")
writeRaster(future_suit, "data/zonation/data/future_suit.tif")
writeRaster(pas, "data/zonation/data/protect_area.tif", datatype = "INT1U")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=18, fig.width=12}
# Current layer
fn <- file.path(sdm_dir, sprintf("suit_%s.tif", gsub(" ", "_", species)))
suit <- rast(fn) %>% mean(na.rm = TRUE)/1000

fn <- file.path(sdm_dir, sprintf("future_suit_%s.tif", gsub(" ", "_", species)))
future_suit <- rast(fn) %>% mean(na.rm = TRUE)/1000

# Integrate
suit_ens <- suit * future_suit / 
        (suit * future_suit + (1 - suit) * (1 - future_suit))

# Weighted by the distance to current AOH
dist_to_aoh <- rast(here("data/suitability/dist_to_aoh.tif"))
suit_fut_cal <- future_suit * (1 - dist_to_aoh)

# Negative exponential decay
exp_dist <- dexp(values(dist_to_aoh), rate = 10)
suit_fut_cal2 <- dist_to_aoh; values(suit_fut_cal2) <- exp_dist
suit_fut_cal2 <- future_suit * suit_fut_cal2

# Zonation 5
rankmap_cr <- rast(here("data/zonation/setup_6/output/rankmap.tif"))
rankmap_bl <- rast(here("data/zonation/setup_1/output/rankmap.tif"))

g1 <- ggplot() +
  geom_spatraster(data = suit) +
  scale_fill_whitebox_c(
    name = "Suitability",
    palette = "muted",
    n.breaks = 12,
    guide = guide_legend(reverse = TRUE)
  ) + ggtitle("Saguinus oedipus (Current)\n ") +
  theme_void() +
  theme(title = element_text(size = 16))

g2 <- ggplot() +
  geom_spatraster(data = future_suit) +
  scale_fill_whitebox_c(
    name = "Suitability",
    palette = "muted",
    n.breaks = 12,
    guide = guide_legend(reverse = TRUE)
  ) + ggtitle("Saguinus oedipus (Future)\n ") +
  theme_void() +
  theme(title = element_text(size = 16))

g3 <- ggplot() +
  geom_spatraster(data = suit_fut_cal) +
  scale_fill_whitebox_c(
    name = "Climate resilient",
    palette = "muted",
    n.breaks = 12,
    guide = guide_legend(reverse = TRUE)
  ) + ggtitle("Saguinus oedipus\n(Future * (1 - distance to AOH))") +
  theme_void() +
  theme(title = element_text(size = 16))

g4 <- ggplot() +
  geom_spatraster(data = suit_fut_cal2) +
  scale_fill_whitebox_c(
    name = "Climate resilient",
    palette = "muted",
    n.breaks = 6,
    guide = guide_legend(reverse = TRUE)
  ) + ggtitle("Saguinus oedipus\n(Future * (dexp distance to AOH))") +
  theme_void() +
  theme(title = element_text(size = 16))

g5 <- ggplot() +
  geom_spatraster(data = rankmap_bl) +
  scale_fill_whitebox_c(
    name = "Climate resilient",
    palette = "muted",
    n.breaks = 6,
    guide = guide_legend(reverse = TRUE)
  ) + ggtitle("Saguinus oedipus\n(zonation5, Baseline connectivity)") +
  theme_void() +
  theme(title = element_text(size = 16))

g6 <- ggplot() +
  geom_spatraster(data = rankmap_cr) +
  scale_fill_whitebox_c(
    name = "Climate resilient",
    palette = "muted",
    n.breaks = 6,
    guide = guide_legend(reverse = TRUE)
  ) + ggtitle("Saguinus oedipus\n(zonation5, Interaction connectivity)") +
  theme_void() +
  theme(title = element_text(size = 16))

ggarrange(g1, g2, g3, g4, g5, g6, ncol = 2, nrow = 3)
```

Zonation5 (or our manual calculation) relies heavily on distance-sensitive transform functions (uniform, triangular, Gaussian, and Negative exponential function) to calculate connectivity. Thus, Brain's dispersal distance analysis can be super helpful for us as well.
