---
title: "Niche escape distance and PA connectivity for climatic migration"
author: "Lei Song"
date: "2024-09-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
  body{
  font-size: 16pt;
}
</style>

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(here)
library(sf)
library(terra)
library(rnaturalearth)
library(dplyr)
library(lubridate)
library(megaSDM)
library(tidyterra)
library(ggpubr)
library(stringr)
library(tidyr)
library(ggplot2)
sf_use_s2(FALSE)
select <- dplyr::select

# Set directories
data_dir <- "/Volumes/gazelle"
proj_dir <- "data/niche_connec"
if (!dir.exists(proj_dir)) dir.create(proj_dir)

# Get boundary
bry <- ne_countries(scale = 50, country = "Colombia") %>% 
  select(admin)
# Add a big enough buffer to include any associated land
bry <- bry %>% vect() %>% 
  buffer(width = units::set_units(100000, "km")) %>% 
  st_as_sf()
```

## Prepare environmental variables

Prepare Bioclimatic variables from CHELSA, and land cover map from Chen et al. They have the same extent (buffered Colombia), resolution (10 km), projection (latlong), so can be used directly in any workflow.

```{r, eval=FALSE}
dir_chelsa <- file.path(data_dir, "CHELSA_V2/GLOBAL/climatologies")
env_dir <- file.path(proj_dir, "env")
if (!dir.exists(env_dir)) dir.create(env_dir)
years <- c("1981-2010", "2011-2040", "2041-2070", "2071-2100")
mod <- "GFDL-ESM4"
ssp <- 'ssp370'

# BIO1, BIO2, BIO12, BIO15, BIO10/(BIO10 + BIO11)
vars <- c("bio1", "bio2", "bio12", "bio15", "bio10", "bio11")
envs <- lapply(years, function(yr){
  fname <- file.path(env_dir, sprintf("chelsa_%s.tif", yr))
  if (!file.exists(fname)){
    if (yr == "1981-2010"){
      src_dir <- file.path(dir_chelsa, yr, "bio")
      lyrs <- rast(file.path(src_dir, sprintf("CHELSA_%s_%s_V.2.1.tif", vars, yr)))
    } else {
      src_dir <- file.path(dir_chelsa, yr, mod, ssp, "bio")
      lyrs <- rast(file.path(src_dir, sprintf("CHELSA_%s_%s_%s_%s_V.2.1.tif", 
                                              vars, yr, tolower(mod), ssp)))
    }
    lyrs <- lyrs %>% crop(bry) %>% mask(bry)
    names(lyrs) <- vars
    lyrs$bio120 <- lyrs$bio10 / (lyrs$bio10 + lyrs$bio11)
    lyrs <- subset(lyrs, c(1:4, 7))
    
    # Write out
    writeRaster(lyrs, fname)
  } else {
    lyrs <- rast(fname)
  }
  
  # Return
  wrap(lyrs)
}); names(envs) <- years
```

## Prepare inputs for mesaSDM

I kind of like this package. It is much more friendly than `biomod2`.

```{r}
sdm_dir <- file.path(proj_dir, "sdm")
if(!dir.exists(sdm_dir)) dir.create(sdm_dir)
# Current for training
TSEnv <- TrainStudyEnv(input_TA = unwrap(envs$`1981-2010`),
                       output = sdm_dir)

# Future ones for projection
PredictEnv(studylayers = TSEnv$study,
           futurelayers = lapply(envs[years[-1]], unwrap),
           time_periods = c(2010, 2040, 2070, 2100),
           output = sdm_dir,
           scenario_name = "RCP3.7")
```

## Occurrences

```{r}
occ_dir <- file.path(proj_dir, "occurrences")
if(!dir.exists(occ_dir)) dir.create(occ_dir)

# Giant Anteater, Jaguar, Spectacled Bear, White-tailed deer, Mountain Tapir
extent_occ <- ext(TSEnv$training$bio1) %>% as.vector()
spplist <- c("Saguinus oedipus",
             "Tremarctos ornatus",
             "Odocoileus virginianus",
             "Tapirus pinchaque")

Occurrences <- OccurrenceCollection(spplist = spplist,
                                    output = occ_dir,
                                    trainingarea = extent_occ)
# Update species list
spplist <- Occurrences$Scientific.Name

# First, get the list of the occurrence files
occlist <- list.files(occ_dir, pattern = ".csv", full.names = TRUE)

OccurrenceManagement(occlist = occlist,
                     output = occ_dir,
                     envextract = TRUE,
                     envsample = TRUE,
                     nbins = 25,
                     envdata = TSEnv$training)

# Extract background points
buff_dir <- file.path(sdm_dir, "buffers")

# Generates buffers for each species.
BackgroundBuffers(occlist = occlist,
                  envdata = TSEnv$training,
                  buff_dir,
                  ncores = 6)

# Extract background points
nbg <- 1000

# What proportion of the background points should be sampled from within the buffers?
spatial_weights <- 0.5 

# Should the background points be environmentally subsampled (Varela) or 
# randomly distributed (random)?
sampleMethod <- "Varela" 

bufflist <- list.files(buff_dir, pattern = ".shp$", full.names = TRUE)

# Define the location where the background points will be printed out to (.shp) 
# (This directory will be created if it doesn't already exist)
bg_dir <- file.path(proj_dir, "background")
if(!dir.exists(bg_dir)) dir.create(bg_dir)

BackgroundPoints(spplist = spplist,
                 envdata = TSEnv$training,
                 output = bg_dir,
                 nbg = nbg,
                 spatial_weights = spatial_weights,
                 buffers = bufflist,
                 method = sampleMethod,
                 ncores = 6)
```

## Diagnose the variables

```{r}
envvar <- rep("Bio1,Bio2,Bio12,Bio15,Bio120", length = length(occlist))

# Define a list of the background point files 
# (either created in the BackgroundPoints function or generated separately)
bglist <- list.files(bg_dir, pattern = ".csv", full.names = TRUE)

# In this example, megaSDM overwrites the occurrence and background points,
# but they could be placed in a different folder if requested.
VariableEnv(occlist = occlist,
           bglist = bglist,
           env_vars = envvar,
           occ_output = occ_dir,
           bg_output = bg_dir)
```

## SDM

```{r, eval=FALSE}
# First, define a list of all background and occurrence point files
occlist <- list.files(occ_dir, pattern = ".csv", full.names = TRUE)
bglist <- list.files(bg_dir, pattern = ".csv", full.names = TRUE)

# Define where the results of the MaxEnt model runs will be printed out to (as .lambdas files)
model_dir <- file.path(sdm_dir, "models")
if (!dir.exists(model_dir)) dir.create(model_dir)

# "nrep" is set to 4, meaning that the MaxEnt algorithm will run 4 times with different
# subsets of occurrence points for a better representation of the habitat suitability.
MaxEntModel(occlist = occlist,
            bglist = bglist,
            model_output = model_dir,
            ncores = 6,
            nrep = 4,
            alloutputs = FALSE)
```

## Projection

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# First, create a list of the time periods and climate scenarios used in the analysis
# (starting with the year the model is trained on)
time_periods <- c(2010, 2040, 2070, 2100)
scenarios <- "RCP3.7"
study_dir <- file.path(sdm_dir, "studyarea")
predict_dir <- file.path(sdm_dir, "RCP3.7", time_periods[-1])
result_dir <- file.path(sdm_dir, "results")
if(!dir.exists(result_dir)) dir.create(result_dir)

# Other options are also available (check the documentation page)
MaxEntProj(input = model_dir,
           time_periods = time_periods,
           scenarios = scenarios,
           study_dir = study_dir,
           predict_dirs = list(predict_dir),
           output = result_dir,
           aucval = 0.7,
           ncores = 6)
```

### Time maps

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=18, fig.width=12}
# # average yearly dispersal rate of a species (given in km/year)
# dispersaldata <- data.frame(
#   Species = spplist, 
#   Rate = c(48.92, 1.21, 6.07, 3.37, 0.96, 3.27))
# 
# dispersalRate(result_dir = result_dir,
#               dispersaldata = dispersaldata,
#               time_periods = time_periods,
#               scenarios = scenarios,
#               ncores = 2)
# 
# createTimeMaps(result_dir, 
#                time_periods, 
#                scenarios, 
#                dispersal = TRUE, 
#                dispersaldata = dispersaldata, 
#                ncores = 2)

createTimeMaps(result_dir = result_dir,
               time_periods = time_periods,
               scenarios = scenarios,
               dispersal = FALSE,
               ncores = 6)
```

```{r}
# Construct value table
vals <- rbind(rep(0, length(time_periods)), rep(1, length(time_periods))) %>% 
  data.frame() %>% setNames(as.character(time_periods)) %>% 
  expand(!!!rlang::syms(as.character(time_periods)))
vals$raster_value <- vals$`2010` * 10^(ncol(vals) - 1) +
  vals$`2040` * 10^(ncol(vals) - 2) +
  vals$`2070` * 10^(ncol(vals) - 3) +
  vals$`2100` * 10^(ncol(vals) - 4)

# Transform the table to objective
vals <- vals %>% 
  mutate(class = ifelse(raster_value == 0, "Background", 
                        ifelse(str_detect(raster_value, "0"), "Loss",
                               "Gain"))) %>% 
  mutate(class = ifelse(`2010` == 0 & class == "Loss", "Background", class)) %>% 
  mutate(class = ifelse(raster_value == 1111, "Overlap", class))
```

```{r}
niche_convert <- function(result_dir,
                          transit_table){
  # Get crispy maps
  species_dir <- list.dirs(result_dir, full.names = TRUE, recursive = FALSE)
  timemaps <- lapply(species_dir, function(path){
    list.files(file.path(path, "TimeMapRasters"), pattern = ".grd$", 
               full.names = TRUE)
  }) %>% unlist()
  
  # Reclassify the maps based on transit table
  vals_current <- transit_table %>% filter(class %in% c("Loss", "Overlap")) %>% 
    pull(raster_value)
  vals_future <- transit_table %>% filter(class %in% c("Gain", "Overlap")) %>% 
    pull(raster_value)
  
  for (fname in timemaps){
    dst_dir <- gsub("TimeMapRasters", "TimeMapRastersSimplify", dirname(fname))
    if(!dir.exists(dst_dir)) dir.create(dst_dir)
    
    current_map <- rast(fname)
    current_map[!current_map %in% vals_current] <- NA
    current_map[current_map %in% vals_current] <- 1
    fn_to <- file.path(dst_dir, paste("current", basename(fname), sep = "_"))
    writeRaster(current_map, fn_to, datatype = "INT1U", overwrite = TRUE)
    
    future_map <- rast(fname)
    future_map[!future_map %in% vals_future] <- NA
    future_map[future_map %in% vals_future] <- 1
    fn_to <- file.path(dst_dir, paste("future", basename(fname), sep = "_"))
    writeRaster(future_map, fn_to, datatype = "INT1U", overwrite = TRUE)
  }
}

# Transit Time maps
niche_convert(result_dir = result_dir, transit_table = vals)
```

## Calculate niche escape distance

```{r}
escape_distance <- function(result_dir, normalize = TRUE, weight = TRUE){
  # Get crispy maps
  species_dir <- list.dirs(result_dir, full.names = TRUE, recursive = FALSE)
  timemaps <- lapply(species_dir, function(path){
    list.files(file.path(path, "TimeMapRastersSimplify"), pattern = ".grd$", 
               full.names = TRUE)
  }); names(timemaps) <- basename(species_dir)
  
  for (nm in names(timemaps)){
    maps <- rast(timemaps[[nm]])
    
    # Calculate distance
    target_map <- mask(maps[[1]], maps[[2]], inverse = TRUE, updatevalue = 0)
    
    # TODO: clean up small far away patches
    
    # work for [0, 1], all zero, all one
    dist_map <- distance(target_map, target = 1, unit = "km")
    
    # Normalize
    if (normalize) dist_map <- stretch(dist_map, minv = 0, maxv = 1, maxq = 0.99)
    
    if (weight){
      # Calculate weight
      sizes <- cellSize(target_map, unit = "km")
      area_ints <- sum(maps[[1]], maps[[2]])
      area_ints <- global(mask(sizes, area_ints), "sum", na.rm = TRUE)[[1]]
      area_u <- sum(maps[[1]], maps[[2]], na.rm = TRUE)
      area_u <- global(mask(sizes, area_u), "sum", na.rm = TRUE)[[1]]
      w <- 1 - area_ints / area_u
      
      # Apply weight
      dist_map <- dist_map * w
    }
    
    # Save out
    rcp <- str_extract(timemaps[[nm]][[1]], "RCP[0-9]{1}.[0-9]{1}")
    fn_to <- file.path(result_dir, nm, sprintf("escape_distance_%s.tif", rcp))
    writeRaster(dist_map, fn_to, overwrite = TRUE)
  }
}

escape_distance(result_dir = result_dir)
```

## Calculate importance of pixels to support PA connectivity

```{r}
pas <- st_read("data/NPs/runap2Polygon.shp")
calculate_connectivity <- function(result_dir,
                                   pas,
                                   normalize = TRUE){
  # Get crispy maps
  species_dir <- list.dirs(result_dir, full.names = TRUE, recursive = FALSE)
  timemaps <- lapply(species_dir, function(path){
    c(list.files(file.path(path, "TimeMapRastersSimplify"), pattern = ".grd$", 
               full.names = TRUE),
    list.files(path, pattern = "^escape_distance", full.names = TRUE))
  }); names(timemaps) <- basename(species_dir)
  
  for (nm in names(timemaps)){
    maps <- rast(timemaps[[nm]])
    
    # Get the associated nodes
    extent <- sum(maps[[1]], maps[[2]], na.rm = TRUE)
    nodes <- mask(rasterize(pas, extent), extent)
    
    # Calculate cost distance to nodes
    cost_surface <- mask(1 - maps[[3]], nodes, inverse = TRUE, updatevalue = -1)
    cost_distance <- costDist(cost_surface, target = -1)
    
    if (normalize) cost_distance <- stretch(cost_distance, minv = 0, maxv = 1)
    
    # Save out
    rcp <- str_extract(timemaps[[nm]][[1]], "RCP[0-9]{1}.[0-9]{1}")
    fn_to <- file.path(result_dir, nm, sprintf("cost_distance_%s.tif", rcp))
    writeRaster(cost_distance, fn_to, overwrite = TRUE)
  }
}

calculate_connectivity(result_dir = result_dir, pas = pas)

# Continuous
calculate_connectivity <- function(result_dir,
                                   pas,
                                   normalize = TRUE){
  # Get crispy maps
  species_dir <- list.dirs(result_dir, full.names = TRUE, recursive = FALSE)
  timemaps <- lapply(species_dir, function(path){
    c(list.files(file.path(path, "TimeMapRastersSimplify"), pattern = ".grd$", 
               full.names = TRUE),
    list.files(path, pattern = "^escape_distance", full.names = TRUE),
    file.path(path, "2010_ensembled.grd"),
    file.path(path, "RCP3.7/2040_RCP3.7_ensembled.grd"),
    file.path(path, "RCP3.7/2070_RCP3.7_ensembled.grd"),
    file.path(path, "RCP3.7/2100_RCP3.7_ensembled.grd"))
  }); names(timemaps) <- basename(species_dir)
  
  for (nm in names(timemaps)){
    maps <- rast(timemaps[[nm]])
    
    # Suitability
    suit <- mean(maps[[4:7]], na.rm = TRUE)
    
    # Get the associated nodes
    extent <- sum(maps[[1]], maps[[2]], na.rm = TRUE)
    nodes <- mask(rasterize(pas, extent), extent)
    
    # Calculate cost distance to nodes
    cost_surface <- mask((1 - suit) * (1 - maps[[3]]), 
                         nodes, inverse = TRUE, updatevalue = -1)
    cost_distance <- costDist(cost_surface, target = -1)
    
    if (normalize) cost_distance <- stretch(cost_distance, minv = 0, maxv = 1)
    
    # Save out
    rcp <- str_extract(timemaps[[nm]][[1]], "RCP[0-9]{1}.[0-9]{1}")
    fn_to <- file.path(result_dir, nm, sprintf("cost_distance_c_%s.tif", rcp))
    writeRaster(cost_distance, fn_to, overwrite = TRUE)
  }
}

calculate_connectivity(result_dir = result_dir, pas = pas)
```

